---
title: "Comparative transcriptomic analysis of breast cancer at single cell level"
author: "Alba Machado-Lopez"
date: "December 2022"
output: html_document
knit: (function(input_file, encoding) {
    out_dir <- 'docs';
    rmarkdown::render(input_file,
      encoding=encoding,
      output_file=file.path(dirname(input_file), out_dir, 'comparative_transcriptomics.html'))})
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load libraries

```{r, message=F}
library(miloR)
library(Seurat)
library(SingleCellExperiment)
library(scater)
library(dplyr)
library(ggplot2)
library(patchwork)
theme_set(theme_bw())
```


# Load objects

We will simply read the files we created with the combined seurat objects from human and mouse cells. 

```{r}
hg19_scRNA <- readRDS("data/R_objects/hg19_RNA.rds")
mm10_scRNA <- readRDS("data/R_objects/mm10_RNA.rds")
```

# Add metadata

We will now add metadata columns indicating the subtype of breast cancer and whether the tumor is resistant or sensitive to chemotherapy

```{r}
hg19_scRNA@meta.data <- hg19_scRNA@meta.data %>%
  mutate(subtype=case_when(
    orig.ident=="HBCx-95"| orig.ident=="HBCx-95_CAPAR"~"Triple_negative",
    orig.ident=="HBCx-22"| orig.ident=="HBCx22-TAMR"~"Luminal")) %>% 
  mutate(drug_resistance=case_when(
    grepl("R", orig.ident)==T~"Resistant",
    grepl("R", orig.ident)==F~"Sensitive"))

mm10_scRNA@meta.data <- mm10_scRNA@meta.data %>%
  mutate(subtype=case_when(
    orig.ident=="HBCx-95"| orig.ident=="HBCx-95_CAPAR"~"Triple_negative",
    orig.ident=="HBCx-22"| orig.ident=="HBCx22-TAMR"~"Luminal")) %>% 
  mutate(drug_resistance=case_when(
    grepl("R", orig.ident)==T~"Resistant",
    grepl("R", orig.ident)==F~"Sensitive"))
```


# Convert to SingleCellExperiment objects

```{r}
hg19_scRNA <- as.SingleCellExperiment(hg19_scRNA)
mm10_scRNA <- as.SingleCellExperiment(mm10_scRNA)
```

# Create Milo object

```{r}
# Create object that can store the KNN graph as a reduced dimension
hg19_scRNA <- Milo(hg19_scRNA)
mm10_scRNA <- Milo(mm10_scRNA)
```

# Construct KNN graph

```{r}
hg19_scRNA <- buildGraph(hg19_scRNA, k=20, d=30)
mm10_scRNA <- buildGraph(mm10_scRNA, k=30, d=30)
```

## Define neighbourhoods

```{r}
hg19_scRNA <- makeNhoods(hg19_scRNA, prop = 0.2, k = 20, d=30, refined = TRUE)
plotNhoodSizeHist(hg19_scRNA)

mm10_scRNA <- makeNhoods(mm10_scRNA, prop = 0.2, k = 30, d=30, refined = TRUE)
plotNhoodSizeHist(mm10_scRNA)
```

Note that we have selected k for constructing the graph and defining neighbours to achieve a distribution peak between 50 and 100 

## Count the number of cells in neighbourhoods

```{r}
hg19_scRNA <- countCells(hg19_scRNA, meta.data = data.frame(colData(hg19_scRNA)), sample="orig.ident")

mm10_scRNA <- countCells(mm10_scRNA, meta.data = data.frame(colData(mm10_scRNA)), sample="orig.ident")
```

# Differential abundance testing

In this step we will perform a statistical test to find populations that are significantly more or less abundant in one group in comparison to another. 

- For human cells:

```{r}
# Create design matrix
design_hg19 <- data.frame(colData(hg19_scRNA))[,c("subtype", "drug_resistance","orig.ident")]
design_hg19 <- distinct(design_hg19)
rownames(design_hg19) <- design_hg19[,"orig.ident"]
design_hg19

# Calculate distances
hg19_scRNA <- calcNhoodDistance(hg19_scRNA, d=30)

# Perform the statistical test
comparison_subtype <- testNhoods(hg19_scRNA, design = ~ subtype, design.df = design_hg19)
comparison_resistance <- testNhoods(hg19_scRNA, design = ~ drug_resistance+subtype, design.df = design_hg19)


```

- Same approach for murine cells:

```{r}
# Create design matrix
design_mm10 <- data.frame(colData(mm10_scRNA))[,c("subtype", "drug_resistance","orig.ident")]
design_mm10 <- distinct(design_mm10)
rownames(design_mm10) <- design_mm10[,"orig.ident"]
design_mm10

# Calculate distances
mm10_scRNA <- calcNhoodDistance(mm10_scRNA, d=30)

# Perform the statistical test
comparison_subtype_mm10 <- testNhoods(mm10_scRNA, design = ~ subtype, design.df = design_mm10)
comparison_resistance_mm10 <- testNhoods(mm10_scRNA, design = ~ drug_resistance+subtype, design.df = design_mm10)

```

## Check that the test was balanced

```{r}
ggplot(comparison_subtype, aes(PValue)) + geom_histogram(bins=50)
ggplot(comparison_resistance, aes(PValue)) + geom_histogram(bins=50)
ggplot(comparison_subtype_mm10, aes(PValue)) + geom_histogram(bins=50)
ggplot(comparison_resistance_mm10, aes(PValue)) + geom_histogram(bins=50)
```

## Volcano plots of neighbourhoods

```{r}
ggplot(comparison_subtype, aes(logFC, -log10(SpatialFDR))) +
  geom_point() +
  geom_hline(yintercept = 1)

ggplot(comparison_resistance, aes(logFC, -log10(SpatialFDR))) +
  geom_point() +
  geom_hline(yintercept = 1)

ggplot(comparison_subtype_mm10, aes(logFC, -log10(SpatialFDR))) +
  geom_point() +
  geom_hline(yintercept = 1)

ggplot(comparison_resistance_mm10, aes(logFC, -log10(SpatialFDR))) +
  geom_point() +
  geom_hline(yintercept = 1)
```

```{r}
hg19_scRNA <- buildNhoodGraph(hg19_scRNA)

## Plot single-cell UMAP
umap_hg19 <- plotReducedDim(hg19_scRNA, dimred = "UMAP", colour_by="cluster_name", text_by = "cluster_name", text_size = 3) +
  guides(fill="none")

## Plot neighbourhood graph
nh_graph_hg19 <- plotNhoodGraphDA(hg19_scRNA, comparison_subtype, layout="UMAP",alpha=0.05)
nh_graph_resistance_hg19 <- plotNhoodGraphDA(hg19_scRNA, comparison_resistance, layout="UMAP",alpha=0.05)

umap_hg19 + nh_graph_hg19 + nh_graph_resistance_hg19 +
  plot_layout(guides="collect") + ggtitle("Comparison between subtypes")
```

Now we'll do the same for mouse cells

```{r}
mm10_scRNA <- buildNhoodGraph(mm10_scRNA)

## Plot single-cell UMAP
umap_mm10 <- plotReducedDim(mm10_scRNA, dimred = "UMAP", colour_by="cluster_name", text_by = "cluster_name", text_size = 3) +
  guides(fill="none")

## Plot neighbourhood graph
nh_graph_mm10 <- plotNhoodGraphDA(mm10_scRNA, comparison_subtype, layout="UMAP",alpha=0.05)
nh_graph_resistance_mm10 <- plotNhoodGraphDA(mm10_scRNA, comparison_resistance, layout="UMAP",alpha=0.05)

umap_mm10 + nh_graph_mm10 + nh_graph_resistance_mm10 +
  plot_layout(guides="collect") + ggtitle("Comparison between subtypes")
```
